Railway: Adaptive Disk Storage for Interaction Graphs

### 2014-09-19 docs/edbt2015/text/ilp.tex
ILP formulation is started

\clearpage
\newpage
\section{Partitioning}
This section is still a work-in-progress. We are trying to find the algorithms.

\subsection{Non-Overlapping Attributes}

\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
minimizes the query I/O and bounds the storage cost by some upper limit.}

\subsubsection{Integer Linear Program Formulation}\label{subsubsec:nov-ilp}
We present an ILP formuation of the problem. For this purpose, we define a 
number of binary ($0$ or $1$) variables: 
\begin{itemize}
\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
is in partition $p$, $0$ otherwise.
\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
\end{itemize}

Each of these variables serve a purpose:
\begin{itemize}
\item $x$s define the attribute-to-partition assignment.
\item $y$s help formulate the query I/O contribution
of partitions, excluding their assigned attributes.
\item $z$s help formulate the query I/O contribution
of partitions, only considering their assigned attributes.
\item $u$s help formulate the storage overhead requirement.
\end{itemize}

@@

@@

\begin{figure}[!t]
\begin{mdframed}
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} w(q)\cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \Big) \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}
\end{mdframed}
\caption{ILP formulation for the non-overlapping partitioning}
\label{fig:nov-ilp}
\end{figure}

@@

\subsubsection{Heuristic Solution}

\paragraph*{Approach.$\,$}
TODO

\begin{algorithm}[ht]
\scriptsize
\caption{Algorithm for partitioning blocks into sub-blocks with non-overlapping attributes.}
\label{alg:non-overlappingP}
\KwData{$B$: block, $Q$: set of queries}
$c^*\leftarrow \infty$ \tcp*{Lowest cost over all \# of partitions}
\For(\tcp*[f]{For each possible \# of partitions}){$k=1$ to $|A|$}{
   $R[i]\leftarrow \emptyset, \forall i\in [1..k]$ \tcp*{Initialize partitions}
   \For(\tcp*[f]{For each attribute}){$a \in A$\textnormal{, in decr.\/ order of }$f(a)$}{
      $c\leftarrow \infty$ \tcp*{Lowest cost over all assignments}  
      $j\leftarrow -1$ \tcp*{Best partition assignment}
      \For(\tcp*[f]{For each partition assignment}){$i\in [1..k]$} {
         $R[i]\leftarrow R[i] \cup \{a\}$\tcp*{Assign attribute}
         \If(\tcp*[f]{If query cost is lower}){$L(R, B, Q)<c$}{
            $c\leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
            $j\leftarrow i$\tcp*{Update the best partition}
         }
         $R[i]\leftarrow R[i] \setminus \{a\}$\tcp*{Un-assign attribute}
      }
      $R[j]\leftarrow R[j] \cup \{a\}$\tcp*{Assign to best partition}
   }
   \lIf(\tcp*[f]{If solution infeasible}){$H(R, B, Q)>\alpha$}{\textbf{break}}
   \If(\tcp*[f]{If solution has lower cost}){$L(R, B, Q)<c^*$}{
     $c^* \leftarrow L(R, B, Q)$\tcp*{Update the lowest cost}
     $\mathcal{P}(B)\leftarrow R$\tcp*{Update the best partitioning}
   } 
}
\Return $\mathcal{P}(B)$ \tcp*{Final set of sub-blocks}
\end{algorithm}

\clearpage
\newpage
\subsection{Overlapping Attributes}

@@

\subsubsection{Integer Linear Program Formulation}
We present an ILP formuation of the problem, as we did for the case of
non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
set of variables and the same objective function. However, the formulation of
the constraints differ.

@@

\begin{figure}[!t]
\begin{mdframed}
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} w(q)\cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \Big) \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \geq 1\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && x_{a,p} - z_{a,p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
\sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_e(B)\cdot x_{a,p}  \Big)\leq s(B)\cdot (1+\alpha)\nonumber
\end{eqnarray}
\end{mdframed}
\caption{ILP formulation for the overlapping partitioning}
\label{fig:ov-ilp}
\end{figure}

@@

\subsubsection{Heuristic Solution}

\paragraph*{Approach.$\,$} We start the algorithm with a   partitioning based
on what queries we have seen. Every query gets its own sub-block. This is the
``ideal'' partitioning, because the I/O cost would be minimized for every
query that we would have seen. The algorithm iteratively combines the two
partitions that are closest together.  After each combination of partitions,
the algorithm calculate the storage overhead for the partitioning. The
algorithm stops when the  storage cost is below some specified threshold.  The
result is the block partitioning.

\begin{algorithm}[ht]
\scriptsize
\caption{Algorithm for partitioning blocks into sub-blocks with overlapping attributes.}
\label{alg:overlappingP}
\KwData{$B$: block, $Q$: set of queries}
$\mathcal{P}(B) \leftarrow \{q.A: q\in Q\}$ \tcp*{Each query gets its own sub-block}
\While(\tcp*[f]{Until storage overhead is below $\alpha$}){$H(\mathcal{P},B) > \alpha$}{
  $c^{*}\gets \infty $ \tcp*{Lowest cost over all sub-block pairs}
  $(b_x,b_y)\gets (\emptyset,\emptyset)$ \tcp*{Sub-block pair with the lowest cost}
  \For(\tcp*[f]{For each pair of blocks}){$\{b_i,b_j\}\in\mathcal{P}(B)$}{
    $\mathcal{P'}(B) \leftarrow \mathcal{P}(B) \setminus \{b_i, b_j\} \cup \{b_i \cup b_j\}$\\
    $c\gets \frac{L(\mathcal{P}',B,Q)-L(\mathcal{P},B,Q)}{H(\mathcal{P},B)-H(\mathcal{P}',B)}$\tcp*{Cost of merge}
    \If(\tcp*[f]{Cost is lower}){$c<c^{*}$}{
        $c^{*}\gets c$\tcp*{Update the lowest cost}
        $(b_x,b_y)\gets (b_i,b_j)$\tcp*{Update the best pair}
    }
  }
  $\mathcal{P}(B) \leftarrow \mathcal{P}(B) \setminus \{b_x, b_y\} \cup \{b_x \cup b_y\}$ \\
}
\Return $ \mathcal{P}(B)$  \tcp*{Final set of sub-blocks}
\end{algorithm}

>>>>

\section{ILP Solution}\label{sec:ilp}
\noindent
In this section, we formulate the optimal railway design problem as a mixed
Integer Linear Program (ILP). The main challenge is to represent the objective
function and the constraint as a linear combination of potentially integer
variables. We separate the cases of non-overlapping and overlapping
partitioning, as the former case can be formulated using smaller number of 
constraints.

For the ILP formulation, we define a number of binary ($0$ or $1$) variables: 
\begin{itemize}
\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
is in partition $p$, $0$ otherwise.
\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
\end{itemize}

Each of these variables serve a purpose:
\begin{itemize}
\item $x$s define the attribute-to-partition assignments.
\item $y$s help formulate the query I/O contribution of each partition due to
the graph structure they contain (excluding their assigned attributes).
\item $z$s help formulate the query I/O contribution of each partition, only
considering the attributes they are assigned.
\item $u$s help formulate the storage overhead requirement.
\end{itemize}

Here, we assume that the maximum number of partitions is fixed. In fact, we
cannot have more partitions than attributes, so the number of partitions is
upper bounded by $|A|$, and thus $0\leq p<|A|$. However, some of these
partitions can be empty in the optimal solution, which means that the number
of partitions found by the ILP solution is typically lower than the maximum
possible. A simple post-processing step removes empty partitions and creates
the final partitioning to be used.

@@

\subsection{Non-Overlapping Attributes}\label{subsubsec:nov-ilp}

@@

\begin{figure}[!t]
%\begin{mdframed}
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} w(q)\cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \Big) \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p \leq 1 + \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}
%\end{mdframed}
\caption{ILP formulation for the non-overlapping partitioning}
\label{fig:nov-ilp}
\end{figure}

@@

\subsection{Overlapping Attributes}\label{subsubsec:ov-ilp}

@@

We present an ILP formuation of the problem, as we did for the case of
non-overlapping partitions in Section~\ref{subsubsec:nov-ilp}. We use the same
set of variables and the same objective function. However, the formulation of
the constraints differ.

@@

\begin{figure}[!t]
%\begin{mdframed}
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} w(q)\cdot \Big(\sum_{p=1}^{k} \!\!&&\!\! (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \Big) \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \geq 1\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && x_{a,p} - z_{a,p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
\sum_{p=1}^{k} \Big((16\cdot c_e(B) &+& 12 \cdot c_n(B)) \cdot u_p  \nonumber \\ 
+ \sum_{a\in A} s(a) \!\!&\cdot&\!\! c_e(B)\cdot x_{a,p}  \Big)\leq s(B)\cdot (1+\alpha)\nonumber
\end{eqnarray}
%\end{mdframed}
\caption{ILP formulation for the overlapping partitioning}
\label{fig:ov-ilp}
\end{figure}

@@
